Austral Spring Extratropical Southern Hemisphere zonally assymetric circulation \thanks{The research was supported by UBACyT20020170100428BA and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01. Elio Campitelli was supported by a PhD grant from CONICET, Argentina.}

### 2021-07-29 analysis/paper/shceof.tex
Render

One typical way of describing the zonally asymmetric circulation is by the amplitude and phase of zonal waves obtained by Fourier decomposition of geopotential height at each latitude \citep[e.g.][]{vanloon1972, trenberth1980, turner2017}.
This approach suggests that zonal waves 1 and 3 explained most of the variance of the tropospheric extratropical Southern Hemisphere circulation (\citet{vanloon1972}).
However, this methodology rely on the assumption that the circulation can be meaningfully understood in terms of zonal waves of constant amplitude along a latitude circle.
This is not valid for meridionally propagating waves or zonal waves with localised amplitudes.
In the case of the wave 3, for example, \citet{trenberth1985} observed that it played a role in blocking events, but this was mostly due to increased amplitude of a longitudinally localised wave train instead of an hemispheric-scale zonal wave 3.

Another approach to characterising the Southern Hemisphere circulation is by using Empirical Orthogonal Functions (EOF, also known as Principal Component Analysis).
Within the EOF framework, the Southern Annular Mode (SAM) appears as the leading mode of variability of the tropospheric Southern Hemisphere circulation \citep{fogt2020} followed by the two Pacific--South American Patterns (PSA) \citep{mo2001}.
The SAM represents a relatively zonally symmetric pattern of alternating low pressures in polar latitude and a ring of high pressures in high latitudes with an embedded wave 3 pattern that is more prominent in the Pacific sector.
The PSA1 and PSA2 describe the two out-of-phase sides of a meridionally propagating wave train that originates in the Eastern equatorial Pacific and travels towards the South Atlantic following a great-circle arch along the Antarctic Peninsula.
These patterns are derived by applying EOF to temporal anomalies, but \citet{raphael2003} applied EOF methods specifically to zonal anomalies.

EOFs are more flexible than Fourier descomposition modes in the sense that they can capture oscillation patterns that cannot be characterised by purely sinusoidal waves with constant amplitude.
Nonetheless, they are restricted to standing oscillation modes and could not represent properly propagating or phase-varying modes such as zonal waves.
A single EOF can also represent a mixture of two or more physical modes.
The inability of a single EOF to describe travelling waves is what forces the PSA pattern to be described by two EOFs.
\citet{irving2016} characterises the Pacific--South American Pattern in a Fourier framework.
Being a meridionally propagating mode, the PSA cannot be correctly characterised by Fourier decomposition at each latitude circle, so they reprojected meridional wind fields so that the path of the PSA laid on the equator.
On this new projection, they could identify the PSA using Fourier decomposition.

Finally, another, hybrid, methodology commonly used consist on identifying particular features of interest and creating indices using simple methods such as averages and differences.
The Marshal SAM Index \citep{marshall2003}, for example, tries to characterise the SAM by the average difference of sea level pressure between 40ºS and 65ºS, following \citet{gong1999}.
\citet{raphael2004} used a similar methodology to describe the wave 3 in the Southern Hemisphere.
Instead of using Fourier to compute the amplitude and phase of the zonal wave, they averaged standardised geopotential height anomalies in three points representing the location of the ridges of the zonal wave 3 of the climatological mean.
This is equivalent to an index of the amplitude of the zonal wave 3 that projects into the phase of the zonal wave 3 of the mean field.
Similarly, \citet{hobbs2010} propose that the zonally asymmetric circulation in the Southern Hemisphere can be described by the strength and location of two anticyclones located in the sub-Antarctic western hemisphere.
These methods are grounded on other methods to identify the centers of action for the described phenomena and can be useful to characterise features that are not readily apparent with Fourier or EOF methods.
The use of data of only specific regions or locations makes them suitable for easy computation and for extending into less quality data from the past, but also makes them not very robust and could not capture non-stationary behaviours of the patterns.

\textcolor{red}{Falta un parráfo que deje más claro por qué es necesario considerar un enfoque alternativo y cuál va a ser el beneficio de ello. También agregar qué se estudiara conjuntamente la troposfera y la estratósfera }

\textcolor{red}{Falta el objetivo que justifica el uso de la metodología}

In this paper we propose using complex Empirical Orthogonal Functions \citep{horel1984} as a robust extension of EOF analysis that is capable of describing rich structures with meridional and zonal propagation, and amplitude modulation such as the PSA.
We apply this method to characterise the Springtime zonally asymmetric circulation of the extratopical Southern Hemisphere.

@@

The cEOF methodology is applied to SON geopotential height zonal anomalies south of 20\degree S at 50 and 200 hPa, representing a stratsopheric and upper troposhperic level respectively.
Table \ref{tab:corr-ceof-splitted} shows the coefficient of determination between time series of the amplitude of each complex EOF (cEOF) across levels.
There's a high degree of correlation between the respective cEOF1 and cEOF2 at each level.
The spatial patterns of the 50 hPa and 200 hPa cEOFs are also similar (not shown).

@@

The result obtained here is similar to the one obtained by \citet{irving2016}.
From their Fourier decomposition of reprojected meridional wind fields \textcolor{red}{¿En qué región?}, they show that the phase distribution is bimodal and isolate the PSA pattern from the rest of the PSA-like variability by selecting events that are near the peaks of the distribution (compare our Figure~\ref{fig:phase-histogram} with their Figure 6).

@@

Figure \ref{fig:t-vertical} shows the vertical distribution of the regression coefficients on cEOF1 from zonal anomalies of air temperature and zonal anomalies of ozone mixing ratio averaged between 75°S and 45°S.
\textcolor{red}{Deberias aclarar qué significa la terminología que estas utilizando. Lo podes hacer en los métodos y ya queda para todo el texto}.
Temperature zonal anomalies associated with cEOF1 show a clear wave 1 pattern for both real and imaginary components throughout the atmosphere above 250 hPa with a change in sign above 10 hPa.
Following hydrostatic balance, this is the level in which the geopotential anomaly have maximum amplitude (not shown).

@@

\textcolor{red}{Revisa los labels de la colorbar para la figura 13. Además ojo que solamente sirven para la columna a, faltan los correspondientes a la streamfunction y el vector de referencia para wave flux}

@@

\textcolor{red}{Al igual que en la introducción, te falta establecer más claro el objetivo, después iría lo que escribiste.}

We studied the two leading modes of co-variability of September-October-November tropospheric and stratospheric zonal anomalies by applying complex Empirical Orthogonal Functions.

@@

We further show that the Imaginary cEOF2 is closely related to the Southern Annular Mode in the troposphere.
In fact, it has a closer resemblance of the zonally asymmetric portion of the SAM.

This raises the possibility that the asymmetric component of the SAM is actually a statistical contamination of the PSA mode.

@@

On the other hand, over Australia, both the Real and Imaginary phase are associated with significant precipitation anomalies, and we further show that the direction of maximum impact is not aligned with our chosen rotation of cEOF2.
However, this underscore the benefit of using complex EOF, since it would trivial to rotate it.

\textcolor{red}{Debería cerrarse con algún parráfo que diga qué tiene de bueno trabajar con esta metodología y qué permitiría hacer a futuro (se puede poner que sería deseable extenderlo a otras estaciones, etc.)}

>>>>

For example, \citet{raphael2007} suggests that variability in the planetary wave 3 projected onto its climatological location is associated with Antarctic sea ice concentration anomalies. \citet{irving2015} showed that planetary wave amplitude in general is associated sea ice concentration, as well as temperature and precipitation anomalies. Another characteristic of the zonally asymmetric circulation are the Pacific--South American Patterns \citep{mo2001}, which are associated with precipitation anomalies in South America.

Some of these patterns of variability appear to have experienced secular changes. \citet{raphael2003} found that the amplitude of the zonal wave 1 experienced a large increase and that the zonal wave 3 experienced changes in its annual cycle between 1958 and 1996.

One typical way of describing the zonally asymmetric circulation is by the amplitude and phase of zonal waves obtained by Fourier decomposition of geopotential height at each latitude \citep[e.g.][]{vanloon1972, trenberth1980, turner2017}.
This approach suggests that zonal waves 1 and 3 explained most of the variance of the tropospheric extratropical Southern Hemisphere circulation (\citet{vanloon1972}).
However, this methodology relies on the assumption that the circulation can be meaningfully understood in terms of zonal waves of constant amplitude along a latitude circle.
This is not valid for meridionally propagating waves or zonal waves with localised amplitudes.
In the case of the wave 3, for example, \citet{trenberth1985} observed that it played a role in blocking events, but this was mostly due to increased amplitude of a longitudinally localised wave train instead of an hemispheric-scale zonal wave 3.

The fourier technique can be generalised to integrate all planetary wave intensity irregardless of wave number by computing the wave envelope as done by \citet{irving2015}. This makes it possible to represent planetary waves with different amplitude at different longitudes, but it removes all information about phase and wave number.

Another approach to characterising the Southern Hemisphere circulation is by using Empirical Orthogonal Functions (EOF, also known as Principal Component Analysis).
Within the EOF framework, the Southern Annular Mode (SAM) appears as the leading mode of variability of the tropospheric Southern Hemisphere circulation \citep{fogt2020} followed by the two PSA \citep{mo2001}.
The SAM represents a relatively zonally symmetric pattern of alternating low pressures in polar latitude and a ring of high pressures in high latitudes with an embedded wave 3 pattern that is more prominent in the Pacific sector.
The PSA1 and PSA2 describe the two out-of-phase sides of a meridionally propagating wave train that originates in the Eastern equatorial Pacific and travels towards the South Atlantic following a great-circle arch along the Antarctic Peninsula.
These patterns are derived by applying EOF to temporal anomalies, but \citet{raphael2003} applied EOF methods specifically to zonal anomalies.

EOFs are more flexible than Fourier decomposition modes in the sense that they can capture oscillation patterns that cannot be characterised by purely sinusoidal waves with constant amplitude.
Nonetheless, they are restricted to standing oscillation modes and could not represent properly propagating or phase-varying modes such as zonal waves.
A single EOF can also represent a mixture of two or more physical modes.
The inability of a single EOF to describe travelling waves is what forces the PSA pattern to be described by two EOFs.
\citet{irving2016} characterises the Pacific--South American Pattern in a Fourier framework.
Being a meridionally propagating mode, the PSA cannot be correctly characterised by Fourier decomposition at each latitude circle, so they reprojected meridional wind fields so that the path of the PSA laid on the equator.
On this new projection, they could identify the PSA using Fourier decomposition.

Finally, another, hybrid, methodology commonly used consist on identifying particular features of interest and creating indices using simple methods such as averages and differences.
The Marshal SAM Index \citep{marshall2003}, for example, tries to characterise the SAM by the average difference of sea level pressure between 40ºS and 65ºS, following \citet{gong1999}.
\citet{raphael2004} used a similar methodology to describe the wave 3 in the Southern Hemisphere.
Instead of using Fourier to compute the amplitude and phase of the zonal wave, they averaged standardised geopotential height anomalies in three points representing the location of the ridges of the zonal wave 3 of the climatological mean.
This is equivalent to an index of the amplitude of the zonal wave 3 that projects into the phase of the zonal wave 3 of the mean field.
Similarly, \citet{hobbs2010} propose that the zonally asymmetric circulation in the Southern Hemisphere can be described by the strength and location of two anticyclones located in the sub-Antarctic western hemisphere.
These methods are grounded on other methods to identify the centres of action for the described phenomena and can be useful to characterise features that are not readily apparent with Fourier or EOF methods.
The use of data of only specific regions or locations makes them suitable for easy computation and for extending back into the data-poor past, but also makes them not very robust and unable to capture non-stationary patterns.

Our objective is to create an index that is able to capture the phase-varying nature of planetary waves and is flexible enough to describe planetary waves with non-constant amplitude in each latitude circle.
We use complex Empirical Orthogonal Functions \citep{horel1984} as a robust extension of EOF analysis that is capable of describing rich structures with meridional and zonal propagation, and amplitude modulation such as the PSA.

We apply this method to characterise the Springtime zonally asymmetric circulation of the extratropical Southern Hemisphere.
To take vertical distribution into account, we analyse both the troposphere and the stratosphere.
We focus the analysis to the September-October-November because that's the trimester in which tropical forcing of zonal waves is maximised (cita).

@@

The cEOF methodology is applied to SON geopotential height zonal anomalies south of 20\degree S at 50 and 200 hPa, representing a stratsopheric and upper troposhperic level respectively.
Table \ref{tab:corr-ceof-splitted} shows the coefficient of determination between time series of the amplitude of each complex EOF (cEOF) across levels.
There's a high degree of correlation between the magnitude of the respective cEOF1 and cEOF2 at each level.
The spatial patterns of the 50 hPa and 200 hPa cEOFs are also similar (not shown).

@@

The result obtained here is similar to the one obtained by \citet{irving2016}.
They show that the phase distribution of the PSA-like variability they detected is bimodal and isolate the PSA pattern from the rest of the PSA-like variability by selecting events that are near the peaks of the distribution (compare our Figure~\ref{fig:phase-histogram} with their Figure 6).

@@

Figure \ref{fig:t-vertical} shows the vertical distribution of the regression coefficients on cEOF1 from zonal anomalies of air temperature and zonal anomalies of ozone mixing ratio averaged between 75°S and 45°S.
Temperature zonal anomalies associated with cEOF1 show a clear wave 1 pattern for both real and imaginary components throughout the atmosphere above 250 hPa with a change in sign above 10 hPa.
Following hydrostatic balance, this is the level in which the geopotential anomaly have maximum amplitude (not shown).

@@

@@

In this study we derived two complex indices of the zonally asymmetric circulation of the September-October-November southern extratropics using Complex Empirical Orthogonal Functions.
This allowed us to characterise both the amplitude and phase of planetary waves that are not perfectly organised as sinusoidal waves with constant amplitude at each latitude circle.

@@

We further show that the Imaginary cEOF2 is closely related to the Southern Annular Mode in the troposphere.
In fact, it has a very close resemblance to the zonally asymmetric portion of the SAM.
This raises the possibility that the asymmetric component of the SAM is actually a statistical contamination of the PSA mode.

@@

On the other hand, over Australia, both the Real and Imaginary phase are associated with significant precipitation anomalies, and we further show that the direction of maximum impact is not aligned with our chosen rotation of cEOF2.
However, this underscores the benefit of using complex EOF, since it would trivial to rotate it.

To the extent that the cEOF2 can be identified with the PSA, what is the advantage of our technique?
The spatial fields that the cEOF2 index describes are in quadrature by construction, which makes it possible to derive a proper amplitude and phase.
Furthermore, because of the way they are constructed, they are not forced to be orthogonal to the SAM.
This makes it possible to study the relationship between the two modes free of the statistical contamination that comes with the usual definition of the PSA as the second and third EOFs.

Further work should extend this analysis to other seasons and further study the relationship between the cEOF2 and the SAM.
In particular, it is not clear whether the close identification between the Imaginary cEOF2 and the Asymmetric SAM arises from physical processes or statistical contamination.
